---
title: "Project Document"
author: "Asif Irfanullah Masum & Ifrat Zaman"
date: "04/13/2023"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```



## Install and load the necessary packages
```{r message=FALSE}
#install.packages(caTools)
#install.packages(dplyr)
#install.packages(nnet)
#install.packages(e1071)
#install.packages(ggplot2)
library(caTools)
library(dplyr)
library(nnet)
library(e1071)
library(ggplot2)
```

## Data Preprocessing and Wrangling

### Loading Data and Exploration 

```{r}
crash_dataset <- read.csv("project_dataset.csv", header = TRUE)
  
# Display the dimensions of the Crash Dataset
cat("Dimensions for Crash Dataset: ", paste(dim(crash_dataset), collapse = " x "))

# Display the structure of the Crash Dataset
cat("\nStructure for Crash Dataset:\n")
str(crash_dataset)

```

### Dividing Dataset into Natural Factor and Human Factor Subset

```{r}
natural_factor_crash <- subset(crash_dataset, select = c(injy_svty_cd, wthr_cd, lit_cd, rd_cond_cd, mdot_regn_cd))
human_factor_crash <- subset(crash_dataset, select = c(injy_svty_cd, vehc_dfct_cd, rstr_not_used_fail, hzrd_actn_cd, alch_susp_ind, drug_susp_ind))

# Display the dimensions of the Natural Factor Subset
cat("Dimensions for Natural Factor Subset : ", paste(dim(natural_factor_crash), collapse = " x "), "\n")

# Display the structure of the Natural Factor Subset
cat("Structure for Natural Factor Subset :\n")
str(natural_factor_crash)

# Display the dimensions of the Human Factor Subset
cat("Dimensions for Human Factor Subset : ", paste(dim(human_factor_crash), collapse = " x "), "\n")

# Display the structure of the Human Factor Subset
cat("Structure for Human Factor Subset :\n")
str(human_factor_crash)
```

### Data Cleaning by removing duplicate and NULL values for Natural Factor Subset

```{r}
cat("Total number of NULL values in Natural Factor Subset : ", paste(sum(is.na(natural_factor_crash))), "\n")
cat("Total number of duplicate values Natural Factor Subset : ", paste(sum(duplicated(natural_factor_crash))), "\n")

natural_factor_crash <- na.omit(natural_factor_crash)
natural_factor_crash <- unique(natural_factor_crash)

cat("Dimensions for Natural Factor Subset after data cleaning: ", paste(dim(natural_factor_crash), collapse = " x "), "\n")
```

### Data Cleaning by removing duplicate and NULL values for Human Factor Subset

```{r}
cat("Total number of NULL values in Human Factor Subset: ", sum(is.na(human_factor_crash)), "\n")
cat("Total number of duplicate values in Human Factor Subset: ", sum(duplicated(human_factor_crash)), "\n")


human_factor_crash <- na.omit(human_factor_crash)
human_factor_crash <- unique(human_factor_crash)

cat("Dimensions for Human Factor Subset after data cleaning: ", paste(dim(human_factor_crash), collapse = " x "), "\n")
```

## Data Visualization for Natural Factors Subset

### Bar plot of Injury Severity Code

```{r}
ggplot(natural_factor_crash, aes(x = injy_svty_cd)) +
  geom_bar(color = "cyan", fill = "cyan") +
  labs(title = "Bar plot of Injury Severity Code")
```

### Bar plot of Weather Code

```{r}
ggplot(natural_factor_crash, aes(x = wthr_cd)) +
  geom_bar(color = "lightblue", fill = "lightblue") +
  labs(title = "Bar plot of Weather Code")
```

### Bar plot of Lighting Code

```{r}
ggplot(natural_factor_crash, aes(x = lit_cd)) +
  geom_bar(color = "lightgreen", fill = "lightgreen") +
  labs(title = "Bar plot of Lighting Code")
```

### Bar plot of Road Condition Code

```{r}
ggplot(natural_factor_crash, aes(x = rd_cond_cd)) +
  geom_bar(color = "black", fill = "black") +
  labs(title = "Bar plot of Road Condition Code")
```

### Bar plot of MDOT Region Code

```{r}
ggplot(natural_factor_crash, aes(x = mdot_regn_cd)) +
  geom_bar(color = "yellow", fill = "yellow") +
  labs(title = "Bar plot of MDOT Region Code")
```

### Scatter plot of Weather Code vs. Injury Severity Code

```{r}
  ggplot(data = natural_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd , y = wthr_cd), color = "orange", position="jitter") +
  labs(title = "Scatter plot of Weather Code Code vs. Injury Severity Code")
```

### Scatter plot of Lighting Code vs. Injury Severity Code

```{r}
  ggplot(data = natural_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd, y = lit_cd), color = "orange", position="jitter") +
  labs(title = "Scatter plot of Lighting Code vs. Injury Severity Code")
```

### Scatter plot of Road Condition Code vs. Injury Severity Code

```{r}
  ggplot(data = natural_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd, y = rd_cond_cd), color = "orange", position="jitter") +
  labs(title = "Scatter plot of Road Condition Code vs. Injury Severity Code")
```

### Scatter plot of MDOT Region Code vs. Injury Severity Code

```{r}
  ggplot(data = natural_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd, y = mdot_regn_cd), color = "orange", position="jitter") +
  labs(title = "Scatter plot of MDOT Region Code vs. Injury Severity Code")
```

## Data Visualization for Human Factors Subset

### Bar plot of Injury Severity Code

```{r}
ggplot(human_factor_crash, aes(x = injy_svty_cd)) +
  geom_bar(color = "cyan", fill = "cyan") +
  labs(title = "Bar plot of Injury Severity Code")
```

### Bar plot of Vehicle Defect Code

```{r}
ggplot(human_factor_crash, aes(x = vehc_dfct_cd)) +
  geom_bar(color = "lightblue", fill = "lightblue") +
  labs(title = "Bar plot of Vehicle Defect Code")
```

### Bar plot of Restraint Usage Code

```{r}
ggplot(human_factor_crash, aes(x = rstr_not_used_fail)) +
  geom_bar(color = "lightgreen", fill = "lightgreen") +
  labs(title = "Bar plot of Restraint Usage Code")
```

### Bar plot of Hazardous Action Code

```{r}
ggplot(human_factor_crash, aes(x = hzrd_actn_cd)) +
  geom_bar(color = "black", fill = "black") +
  labs(title = "Bar plot of Hazardous Action Code")
```

### Bar plot of Alcohol Usage Code

```{r}
ggplot(human_factor_crash, aes(x = alch_susp_ind)) +
  geom_bar(color = "yellow", fill = "yellow") +
  labs(title = "Bar plot of Alcohol Usage Code")
```

### Bar plot of Drugs Usage Code

```{r}
ggplot(human_factor_crash, aes(x = drug_susp_ind)) +
  geom_bar(color = "red", fill = "red") +
  labs(title = "Bar plot of Drugs Usage Code")
```

### Scatter plot of Vehicle Defect Code vs. Injury Severity Code

```{r}
  ggplot(data = human_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd , y = vehc_dfct_cd), color = "lightblue", position="jitter") +
  labs(title = "Scatter plot of Vehicle Defect Code Code vs. Injury Severity Code")
```

### Scatter plot of Restraint Usage Code vs. Injury Severity Code

```{r}
  ggplot(data = human_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd, y = rstr_not_used_fail), color = "lightblue", position="jitter") +
  labs(title = "Scatter plot of Restraint Usage Code vs. Injury Severity Code")
```

### Scatter plot of Hazardous Action Code vs. Injury Severity Code

```{r}
  ggplot(data = human_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd, y = hzrd_actn_cd), color = "lightblue", position="jitter") +
  labs(title = "Scatter plot of Hazardous Action Code vs. Injury Severity Code")
```

### Scatter plot of Alcohol Usage Code vs. Injury Severity Code

```{r}
  ggplot(data = human_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd, y = alch_susp_ind), color = "lightblue", position="jitter") +
  labs(title = "Scatter plot of Alcohol Usage Code vs. Injury Severity Code")
```

### Scatter plot of Drugs Usage Code vs. Injury Severity Code

```{r}
  ggplot(data = human_factor_crash) +
  geom_point(mapping = aes(x = injy_svty_cd, y = drug_susp_ind), color = "lightblue", position="jitter") +
  labs(title = "Scatter plot of Drugs Usage Code vs. Injury Severity Code")
```

## Spliting Data for Training set and Testing set for Natural Factor Subset

```{r}
set.seed(123)
split <- sample.split(natural_factor_crash$injy_svty_cd, SplitRatio = 0.8)
nfcTrainData <- subset(natural_factor_crash, split == TRUE)
nfcTestData <- subset(natural_factor_crash, split == FALSE)
```

## Model Training on Natural Factor Subset

```{r}
# Train a multi-class logistic regression model
logRegModel1 <- multinom(injy_svty_cd ~ ., data = nfcTrainData)

# Make predictions on the test data
predictions1 <- predict(logRegModel1, nfcTestData)

# Check the metrics of the model
cat("Model Accuracy for Human Factor Subset", mean(predictions1 == nfcTestData$injy_svty_cd), "\n")
```

## Spliting Data for Training set and Testing set for Human Factor Subset

```{r}
set.seed(123)
split <- sample.split(human_factor_crash$injy_svty_cd, SplitRatio = 0.8)
hfcTrainData <- subset(human_factor_crash, split == TRUE)
hfcTestData <- subset(human_factor_crash, split == FALSE)
```

## Model Training on Human Factor Subset

```{r}
# Train a multi-class logistic regression model
logRegModel2 <- multinom(injy_svty_cd ~ ., data = hfcTrainData)

# Make predictions on the test data
predictions2 <- predict(logRegModel2, hfcTestData)

# Check the accuracy of the model
cat("Model Accuracy for Human Factor Subset", mean(predictions2 == hfcTestData$injy_svty_cd), "\n")
```
